Zambia Potential Yield Grids

### 2016-02-08 Rmd/demo-figures.Rmd
Many changes, including development of yield-predictors, empirical-yields (GAM fit to DSSAT yields), paper to-do list, etc.

```{r, eval=FALSE}
path <- "external/data/dt/latest/"
il <- input_handler(input_key = "ZA", ybetas = list(1, 1), ybeta_update = 0, 
                    path = path)

@@

zambia <- countriesHigh[countriesHigh@data$SOVEREIGNT == "Zambia", ]
zambia <- spTransform(zambia, CRSobj)
p_figs <- full_path(p_main, "paper/figures/presentation/")
# p_roads <- full_path(p_main, "external/input_devel/roads/tmdpfr_links.sqlite")
load(full_path(p_model, "external/data/parks_roads.rda"))
# roads <- readOGR(p_roads, layer = "tmdpfr_links")
# pas <- raster(full_path(p_bd, "ZA-pas.tif"))
# pmask <- !is.na(pas)
# pmask[pmask == 0] <- NA

>>>>

```{r, eval=FALSE}
setwd("agroEcoTradeoff")
path <- full_path(set_base_path(), "external/data/ZA")
code <- run_code("ZA")
il <- input_handler(path, ybetas = list(1, 1), ybeta_update = 0, code = code)

@@

zambia <- countriesHigh[countriesHigh@data$SOVEREIGNT == "Zambia", ]
zambia <- spTransform(zambia, CRSobj)
p_figs <- full_path(p_main, "paper/figures/presentation/")
# p_roads <- full_path(p_main, "external/input_devel/roads/tmdpfr_links.sqlite")
load(full_path(p_model, "external/data/ZA/parks_roads.rda"))
save(pas, roads, zambia, 
     file = full_path(p_model, "external/data/ZA/parks_roads.rda"))
# roads <- readOGR(p_roads, layer = "tmdpfr_links")
# pas <- raster(full_path(p_bd, "ZA-pas.tif"))
# pmask <- !is.na(pas)
# pmask[pmask == 0] <- NA

### 2016-02-08 Rmd/maize-soy-grid.Rmd
Many changes, including development of yield-predictors, empirical-yields (GAM fit to DSSAT yields), paper to-do list, etc.

Based on simulated potential yields run by Di Tian. 
```{r, eval = FALSE}
library(agroEcoTradeoff)
# library(data.table)
library(lmisc)
# library(dtraster)
# devtools::install_github("PrincetonUniversity/lmisc")

@@

# cellStats(yld_rl$maize$R[["ymu_R0002"]], mean)
# cellStats(yld_rl$soybean$I[["ymu_I990008"]], mean)
plot(yld_rl$soybean$S[["ymu_S990008"]])
plot(yld_rl$soybean$R[["ymu_R990008"]])
plot(yld_rl$maize$R[["y50_R0003"]])
```

## Yield bias correction
### Soybean fix based on GAEZ data

@@

```{r, eval = FALSE} 
cellStats(yld_rl$maize$S[["ymu_S0001"]], mean)
plot(yld_rl$maize$R[["ymu_R0002"]])
plot(yld_rl$maize$S[["ymu_S0002"]])

@@

@@

>>>>

Based on simulated potential yields run by Di Tian. 
```{r, eval = FALSE}
library(agroEcoTradeoff)

@@

# cellStats(yld_rl$maize$R[["ymu_R0002"]], mean)
# cellStats(yld_rl$soybean$I[["ymu_I990008"]], mean)
```

## Yield bias correction
### Approach 1: soybean fix based on GAEZ data

@@

```{r, eval = FALSE} 
# cellStats(yld_rl$maize$S[["ymu_S0001"]], mean)
p_scratch <- fp(p_root, "paper/figures/scratch")
for(i in crops) {  # i <- "maize"
  for(j in c("S", "R", "I")) {  # j <- "S"
    yr <- yld_rl[[i]][[j]]
    yrmu <- calc(yr[[grep("mu", names(yr))]], mean)
    png(fp(p_scratch, paste0(i, "_", j, "_mu.png")), width = 700, height = 700)
    par(mar = c(0, 0, 2, 0))
    plot(yrmu, axes = FALSE, box = FALSE)
    dev.off()
  }
}
# plot(yld_rl$soybean$R[["ymu_R990008"]])
# plot(yld_rl$maize$R[["y50_R0003"]])
# plot(yld_rl$maize$R[["ymu_R0002"]])
# plot(yld_rl$maize$S[["ymu_S0002"]])

@@

### Approach 2: Bias correct against FAO Stats

@@

### Approach 3: Bias-correct against MODIS NPP
```{r, eval = FALSE}
# MODIS NPP
p_npp <- "~/Dropbox/data/npp/mod17a3/"
npp <- lapply(dir(p_npp, full.names = TRUE), function(x) {
  r <- crop(raster(x), yld_rl$maize$S)
  named_out(r, paste0("NPP", gsub("MOD.*.NPP_|.tif", "", basename(x))))
})
names(npp) <- sapply(npp, function(x) names(x))

# Zambia provinces
p_prov <- paste0("/Users/lestes/Dropbox/data/zari/infrastructure/zambia/", 
                 "rda_dump/Final Shape files RDA/")
prov <- readOGR(fp(p_prov, "prov.shp"), layer = "prov")
prov@proj4string <- npp[[1]]@crs
prov$ID <- 1:nrow(prov)

# load in cropland 1 km layer
cropland <- raster(fp(p_cropland, "cropland-1k.tif"))
croplandgcs <- projectRaster(cropland, npp[[1]])
croplandmsk <- croplandgcs > 0  # select pixels having any cropland
provr <- rasterize(prov, croplandmsk, field = "ID")

# select out those pixels from NPP
nppcrops <- lapply(npp, function(x) croplandmsk * x * 0.1)

s <- stack(provr, croplandgcs, stack(npp))
npp_dt <- as.data.table(s)
setnames(npp_dt, c("layer", "cropland.1k"), c("prov", "cland"))
npp_dt <- npp_dt[!is.na(cland) | !is.na(prov), ]
npp_dt <- npp_dt[!is.na(prov), ]
npp_dt <- npp_dt[!is.na(cland), ]
# plot(dt_to_raster(npp_dt[, .(x, y, cland)], prov@proj4string))

# calculate weighted means by province
npp_mu <- npp_dt[, lapply(.SD, function(x) {
  sum(x * 0.1 * cland, na.rm = TRUE) / sum(cland)
}), by = prov, .SDcols = c("cland", names(npp))]
npp_dt[, sum(cland), by = prov]

npp_mu2 <- npp_dt[, lapply(.SD, function(x) {
  weighted.mean(x * 0.1, w = cland, na.rm = TRUE)
}), by = prov, .SDcols = c("cland", names(npp))]  # slightly diff. rounding?

npp_mu <- merge(npp_mu, prov@data[, c("PROVINCE", "ID")], by.x = "prov", 
                by.y = "ID")

# FAO country-stat data, which have problems, it seems (see checks below)
prod <- fao[Crop %in% c("Maize", "Soybeans") & var == "prod", ]
setnames(prod, "Value", "prod")
setkeyv(prod, c("prov", "Year", "Crop"))
prod[, c("var", "provcode", "cropcode", "Flag") := NULL]
pa <- fao[Crop %in% c("Maize", "Soybeans")  & var == "pa", ]
pa[Crop == "Soybeans", hist(pa)]
setnames(pa, "Value", "pa")
setkeyv(pa, c("prov", "Year", "Crop"))
pa[, c("var", "provcode", "cropcode", "Flag") := NULL]

# prodpa <- prod[pa]
# prodpa[, yld := prod / pa]
# prodpa[Crop == "Soybeans" & Year %in% 2007:2009 & prov == "Western"]
# fao[Year == 2011 & Crop == "Soybeans"]

# plot(npp_mu2$NPP2008, npp_mu$NPP2008)
# prodpa[Crop == "Maize" & Year == 2011, ]
# prodpa[Crop == "Maize" & prov == "Southern", 
#        plot(Year, yld, type = "l", ylim = c(0, 6))]
# cols <- rainbow(9)
# for(i in 1:9) {
#   prodpa[Crop == "Maize" & prov == unique(prodpa$prov)[i], 
#          lines(Year, yld, col = cols[i])]
# }
# prodpa[Crop == "Soybeans" & prov == "Southern", 
#        plot(Year, yld, type = "l", ylim = c(0, 6))]
# for(i in 1:9) {
#   prodpa[Crop == "Soybeans" & prov == unique(prodpa$prov)[i], 
#          lines(Year, yld, col = cols[i])]
# }

# Actually forget the above, too unreliable. Use the PHS data I found from ZARI, # cross-checked against FAO country level production and harvested area data 
# (to make PHS and CFS)
p_phs <- "~/Dropbox/data/agricultural/zambia/phs/"
phs <- data.table(readxl::read_excel(fp(p_phs, "phs-1986-2014.xlsx")))
setnames(phs, "YEAR", "Year")

# reshape it
phs_rs <- rbindlist(lapply(names(phs)[-c(1:3)], function(x) {
  DT <- phs[, c("Year", "crop", x), with = FALSE]
  setnames(DT, x, "yld")
  DT[, prov := x]
}))
setkeyv(phs_rs, c("Year", "prov", "crop"))
# phs_rs[crop == "maize", mean(yld, na.rm = TRUE), by = Year][, plot(Year,V1)]
# plot(phs_rs$Year[phs_rs$crop == "maize"], phs_rs$yld[phs_rs$crop == "maize"])

npp_murs <- rbindlist(lapply(names(npp), function(x) { 
  DT <- npp_mu[, c("PROVINCE", x), with = FALSE]
  DT[, Year := gsub("NPP", "", x)]
  setnames(DT, c("PROVINCE", x), c("prov", "NPP"))
  DT
}))
npp_murs[, Year := as.numeric(Year)]
setkeyv(npp_murs, c("Year", "prov"))

# prodmg <- prodpa[npp_murs]
tlm <- lm((a = 1:10) ~ (b = 1:10))
plot(tlm$model[, 2], tlm$model[, 1])
str(tlm)
prodmg <- phs_rs[npp_murs]
mznpp <- prodmg[crop == "maize", lapply(.SD, mean), .SDcols = c("yld", "NPP"),
                by = prov]
prodmg[crop == "soybean", lapply(.SD, mean), .SDcols = c("yld", "NPP"),
       by = prov][, summary(lm(yld ~ NPP))]
prodmg[crop == "soybean", lapply(.SD, mean), .SDcols = c("yld", "NPP"),
       by = prov][, plot(NPP, yld)]

prodmg[crop == "maize", lapply(.SD, mean), .SDcols = c("yld", "NPP"),
                by = prov][,  {
                  plot(yld ~ NPP)
                  npplm = lm(yld ~ NPP)
                  abline(npplm)
                  summary(npplm)
                }]
```

The relationship between mean annual NPP over croplands and maize, both by province, is:

```
Call:
lm(formula = yld ~ NPP)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.85083 -0.10149  0.05117  0.26377  0.40294

Coefficients:
             Estimate Std. Error t value Pr(>|t|)  
(Intercept) -2.369097   1.531678  -1.547   0.1658  
NPP          0.006685   0.002326   2.874   0.0238 *
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.4277 on 7 degrees of freedom
Multiple R-squared:  0.5413,	Adjusted R-squared:  0.4758 
F-statistic: 8.262 on 1 and 7 DF,  p-value: 0.02384
```
Soy on the other hand has no clear relationship, probably because it is a minor crop and covers only a small portion of the area.

Now to correct DSSAT yields using these NPP
```{r, eval=FALSE}
p_bd <- fp(p_root, "external/input_devel/biodiversity")
remfrac <- raster(full_path(p_bd, fname = "ZA-remfrac.tif"))

# calculate mean NPP of relatively undisturbed pixels in neighborhoods 
remmsk <- remfrac > 0.9  # areas having greater > 90% intact

w <- matrix(nrow = 5, ncol = 5, rep(1, 25))
remmsk[remmsk == 0] <- NA

nppmu <- calc(stack(npp) * 0.1, mean)
nppalb <- projectRaster(nppmu, remmsk)
nppalb <- raster::mask(nppalb, remmsk)

# moving window
npp_fmu <- focal(nppalb, w, fun = mean, na.rm = TRUE)
npp_fmx <- focal(nppalb, w, fun = max, na.rm = TRUE)
plot(npp_fmx)
plot(npp_fmu)

# NPP to Z-score
npp_z <- (npp_fmu - cellStats(npp_fmu, mean)) / cellStats(npp_fmu, sd)

dr <- disaggregate(yld_rl[["maize"]][["R"]]$ymu_R0002, fact = 10)
dralb <- projectRaster(dr, npp_fmu)
par(mfrow = c(1, 2), mar = rep(0, 4))
plot(npp_z); plot(dralb)
plot(dralb, npp_fmu)

plot(dr2alb)

dr_z <- (dralb - cellStats(dralb, mean)) / cellStats(dralb, sd)
plot(npp_z)

dr2 <- disaggregate(yld_rl[["soybean"]][["R"]]$ymu_R990008, fact = 10)
dr2alb <- projectRaster(dr2, npp_fmu)
par(mar = rep(0, 4))
plot(dr2alb, npp_fmu)

ydiffs <- lapply(crops, function(x) {
  yr <- lapply(c("S", "R", "I"), function(y) {
    r <- yld_rl[[x]][[y]]
    nms <- grep("mu", names(r))
    rmu <- cellStats(calc(r[[nms]], mean), mean)
    rvar <- r / rmu
    names(rvar) <- names(r)
    list("ratio" = rvar, "mu" = rmu)
  })
  names(yr) <- c("S", "R", "I")
  yr
})
names(ydiffs) <- crops

plot(fpp_mu)
